---
title: "ggplot2: An introduction"
date-format: iso
format: 
  html:
    embed-resources: true
---

## Libraries

We're going to use several new packages today. Install the following via the package manager:

* palmerpenguins
* RColorBrewer
* patchwork


```{r}
#| output: false

library(tidyverse)  # loads ggplot2, dplyr, and others
library(palmerpenguins)  # easy access to the penguins data
library(RColorBrewer) # color palettes
library(patchwork) # laying out sub-figures
```

## References

There are lots of ggplot2 tutorials on the web, but the ggplot2 website and book are definitive resources if you want to understand how to use and extend ggplot2:

* ggplot2 website -- https://ggplot2.tidyverse.org/

* ggplot2 book -- https://ggplot2-book.org/


## Data

We'll be primarily using the now familiar Palmer penguins data set to illustrate ggplot basics.  But first, we'll remove any missing values to simplify our plots:

```{r}
# We're dropping all rows with NA values for illustration purposes
# Don't do this out of habit for your own data though!  
tidy_data <- drop_na(penguins)
```




## Required layers: Data, geometry, and aesthetics

Every ggplot2 figure requires us to define three elements:

1. Data, in the form of a data frame

2. Aesthetics, specifying what gets variables mapped to the various aspects of the geometric mapping 

3. One or more geometric mappings that specify the plot type(s)





## Layer 1 - Data

Remember - ggplot works in layers!

We start with calling ggplot as a function with parenthesis.

The first layer is the data - this gives us an empty canvas to build onto. In this case, we specify our penguins dataset within the parenthesis using `data = penguins`

*Note: You can omit the `data =` part and just call a data frame like `ggplot(tidy_data)`, but it is always good to clearly specify what your code is doing!*

```{r}
#| output: false
ggplot(data = tidy_data)
```

This generated have a blank canvas!


## Layer 2 - Aesthetics = Mapping

In ggplot, 'aesthetics' refers to the mapping of columns of our dataset onto aesthetic attributes of our plot.

In this case, the attributes we want to specify are the x and y coordinates of the plot.

Use the `aes()` function to map the x coordinate to the `species` column in our penguins dataset.

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = body_mass_g))
```

Now we can see that body mass hass appeared on the x-axis of our canvas!

However, there still isn't any data shown. We need to specify a **geometry** that gives our data a shape to display



## Layer 3 - Geometries

Like an oil painting, we are building in layers on top of what we have already done.

In ggplot, we use the + operator after the `ggplot()` call to tell R that the following layer is part of the plot.

Let's start with a basic **density plot**.  A density plot estimates the probability distribution of a variable of interest. The corresponding Geom for generating such a plot is `geom_density`.  .

```{r}
#| output: true
ggplot(data = tidy_data, mapping = aes(x = body_mass_g)) +
  geom_density()
```

You will notice that even though we didn't specify a y-axis variable, the `geom_density` function automatically the y-values of the values we specified in the x-axis call).


## Multiple ways to specify ggplot layers

* Aesthetics specified in a `ggplot` call, are inherited by all geoms. Let's illustrate this by adding a second geom to our density plot:

```{r}
#| output: false
# both geoms inherit the aesthetics
ggplot(tidy_data, 
       aes(x = body_mass_g)) +  
  geom_density() + 
  geom_rug()
```

The `geom_rug` function can be very useful in visualizations like this, since it helps show the marginal distributions within the data.


* Aesthetics can be specified in an independent call to the `aes()` function:

```{r}
#| output: false
# again aesthetics inherited by geoms that follow
ggplot(tidy_data) +
  aes(x = body_mass_g) +
  geom_density() +
  geom_rug()

```


* Aesthetics can be specified in the geoms themselves:

```{r}
#| output: false
# here each geom needs to redefine the aesthetics
ggplot(tidy_data) +
  geom_density(aes(x = body_mass_g)) +
  geom_rug(aes(x = body_mass_g, color = species))
```


## ggplot elements can be assigned to variables

ggplot layers can be assigned to variables for further manipulation:

```{r}
#| output: false
data_layer <- ggplot(tidy_data)
aes_layer <- aes(x = body_mass_g)

my_plot <-
  data_layer +
  aes_layer +
  geom_density() 

my_plot
```

## Bar plot for representing categorical variables 

A density plot is an appropriate depiction of an probability distribution  for continuous variables, but not so for categorical variables.  

For a variable like species, a basic **bar plot** with height corresponding to the count of the data values is more appropriate. The geom for this is `geom_bar()`


```{r}
#| output: false
ggplot(data = tidy_data, mapping = aes(x = species)) +
  geom_bar()
```





## Two variables

Let's add a second variable to our density plots by using color to represent the species column of our data frame:

```{r}
#| output: true
ggplot(data = tidy_data,
       mapping = aes(x = body_mass_g, 
                     color = species)) +
  geom_density() +
  geom_rug()
```



## Box plots

A boxplot is another statistical summary plot that can be useful for depicting multiple sub-distributions simultaneously.


```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = body_mass_g,  
                     y = species)) +
  geom_boxplot()
```

Above I drew the boxplots horizontally to emphasize the parallels to the density plots, but you're probably more used to looking at boxplots like this (note the species and body_mass_g axes are reversed):

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species,  
                     y = body_mass_g)) +
  geom_boxplot() +
  geom_point(alpha = 0.5)
```

Let's add another geom layer to our boxplot so we can look at both raw data and the summary plot together:

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species,  
                     y = body_mass_g)) +
  geom_boxplot() +
  geom_point()
```





### Making things look nicer - transparency

Woo! We have nice boxplots that clearly show where the means of our clusters are different.

However, this plot still looks kind of ugly. Lets add some optional variables to make it nicer to look at and clearly display our data points.

We can do this by first adjusting the opacity of the points Opacity in programming languages is often specified using the value `alpha`. `alpha = 0` is completely transparent while `alpha = 1` is totally opaque; values in between 0 and 1 show intermediate transparency.

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_point(alpha = 0.35) 
```

We can now see the individual points a little better! However, they're still all smushed up against each other in a straight line, and it can be hard to see how the distribution of the points works

### Jittering

We can add some random horizontal spacing and noise to this by adding what is called **jitter**.

*NOTE: This is just random noise, and does not represent properties of the original data!*

ggplot includes two ways to do this: setting the `position` variable to have the value `position = position_jitter()` or replacing the `geom_point` function with a `geom_jitter` function.


```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.35,
             position = position_jitter())
```


`position_jitter` has two important parameters `height` and `width` that specify the amount of vertical and horizontal jitter. If omitted, the values default to 40% of the resolution of the data: this means the jitter values will occupy 80% of the implied bins.  I find that this default is too large for my tastes, so I typically reduce it:

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.35,
             position = position_jitter(width = 0.15,
                                        height = 0))
```

The other change in the  jittersgraphs above was to suppress the plotting of estimated outlier points in the call to `geom_boxplot()`.  Since we're drawing all the points using `geom_point` the plotting of any outliers by boxplot would be redundant and confusing.

Here's the corresponding version using `geom_jitter`:

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(alpha = 0.35,
              width = 0.15,
              height = 0) 
```



*Note: The spread of the jittered points can look pretty chaotic! If you want to make a very pretty publication-worthy graph while avoiding overplotting, consider installing the package `ggbeeswarm` to make a nicely shaped spread of points.*


## Three variables - Color

We have a lot of variables in our dataset, some of which may be correlated with each other. Lets see if there is any association between a penguins body mass and its sex by coloring our points!

Color is an aesthetic! Lets add a new `aes()` function with a `color` variable to our `geom_point` layer, and specify the `sex` column in our dataset:

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(color = sex),
             alpha = 0.35,
             position = position_jitter(width = 0.15,
                                        height = 0))
```

Wow! It looks qualitatively like male penguins might tend to have larger body mass than female penguins across all species.

To quantitate this, lets split our boxplots up by sex as well. We'll move the color aesthetic up into the ggplot mapping so it effects all layers:

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, 
                     y = body_mass_g,
                     color = sex)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(alpha = 0.35,
             position = position_jitter(width = 0.15,
                                        height = 0))
```

That sort of worked - it looks like our groups are separated nicely in terms of the boxplots, but you'll notice that the point themselves didn't separate by sex. Let's fix that.

## Grouping and dodging

If you want to separate out the points so that they go under their boxplots again instead of jittering them, we can use a `position_dodge` in combination with `group`.

`position_dodge` adjusts the horizontal position of the data by using a grouping variable. In this case, we will set our grouping variable to also be sex, and add our `position_dodge` statement to the geom with a width of 0.75.

*Note: position_dodge - like jitter and alpha - is not an aesthetic and goes outside of the aes parenthesis but within the geom parenthesis.*

*In the weeds: You will notice that a position_dodge with a width of 0.75 aligns suspiciously perfectly with our boxplots. This is because boxplots are actually already using position_dodge by default when you add another variable like color!*

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, 
                     y = body_mass_g,
                     color = sex)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(group = sex),
              position = position_dodge(width = 0.75),
              alpha = 0.5)
  
```

We have lost our jitter, but we can add it back!

To add it back while retaining the grouping, we can simultaneously dodge and jitter by using `position_jitterdodge` and changing `width` to `dodge.width`

```{r}
#| output: false
ggplot(data = tidy_data, 
       mapping = aes(x = species, 
                     y = body_mass_g,
                     color = sex)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(group = sex),
             position = position_jitterdodge(dodge.width = 0.75,
                                              jitter.width = 0.25),
             alpha = 0.5)
```

This plot is looking pretty great!




## Labels and titles

The `labs` function is used to specify axis labels, titles, subtitles, etc.


```{r}
ggplot(data = tidy_data, 
       mapping = aes(x = species, 
                     y = body_mass_g,
                     color = sex)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(group = sex),
             position = position_jitterdodge(dodge.width = 0.75,
                                              jitter.width = 0.25),
             alpha = 0.5) +
  labs(x = "Species",
       y = "Body Mass (g)",
       title = "Relationship between body mass and sex by penguin species",
       caption = "Data from palmerpenguins")
```

## Scatter plots

Boxplots are useful for comparing a numerical data column (body mass) to a categorical data column (species). You can use a simple scatter plot to compare two numerical data columns together, and color by a third categorical data type.


```{r}
#| output: false
scatter_example <-
  ggplot(tidy_data, 
       aes(x = body_mass_g, 
           y = flipper_length_mm,
           color = species)) +
  geom_point()

scatter_example
```

```{r}
#| output: false
better_scatter <- 
  scatter_example + 
  labs(x = "Body mass (g)",
       y = "Flipper length (mm)",
       title = "Penguin morphology",
       caption = "Data from palmerpenguins")

better_scatter
```





## Themes

Themes change key aspects of the non-data related aspects of plots.

There are a variety of default themes:

```{r}
#| output: false
better_scatter + theme_classic()
```

```{r}
#| output: false
better_scatter + theme_minimal()
```

## Color palettes

Color palettes can be used to specify the color used in plot.

The RColorBrewer package provides a number of well designed color schemes, but there are other packages that provide color palettes and you can specify your own palette as well.

The pre-definied RColorBrewer palette's can be seen [at this link](https://renenyffenegger.ch/notes/development/languages/R/packages/RColorBrewer/index). You can also view these palettes in your Quarto document as follows:

```{r}
#| fig-width: 6
#| fig-height: 8

# the lines above are special Quarto related syntax for formatting figures
# see the following link about figure options that can be set in quarto docs
# https://quarto.org/docs/reference/formats/html.html#figures

display.brewer.all()
```

### Color and fill

Color and fill are the chief aesthetic properties that are effected by palettes and in geoms that use both, they can be set indepedently.  Color modify functions in ggplot start with `scale_color_` while fill modifying functions start with `scale_fill_`.


If we wish to use one of the RColorBrewer palettes to changes the color of points in a scatter plot, `scale_color_brewer` can be used:

```{r}
#| output: false
# For categorical data I like the "Dark2" color scheme
better_scatter +
  scale_color_brewer(palette = "Dark2") 
```

Here's another one of the RColorBrewer palettes.

```{r}
#| output: false
better_scatter + 
  scale_color_brewer(palette = "Set1") 
```

### Defining your own color palettes

If you want to define your own palette you can do it as so:

```{r}
# colors can be specified in various ways
# ggplot recognizes both hex codes, SVG color names (https://johndecember.com/html/spec/colorsvg.html) 
# In this example we construct a custom palette using color names

custom_colors <- c("firebrick", "steelblue", "goldenrod")


better_scatter + 
  scale_color_manual(values = custom_colors) 

```

### scale_fill_ examples

`geom_density` and `geom_histogram` have both color and fill aesthetics:


```{r}
#| output: false
ggplot(tidy_data, aes(x = body_mass_g,
                     color = species,
                     fill = species)) +
  geom_density(alpha = 0.5) + 
  scale_fill_brewer(palette = "Dark2") + 
  scale_color_brewer(palette = "Dark2")
```

It can be useful to specify both fill and color palettes in a list, and assign that a variable name so you can refer to those settings easily:

```{r}
#| output: false
preferred_palette = "Dark2"

my_colors = list(
  scale_fill_brewer(palette = "Dark2"),
  scale_color_brewer(palette = "Dark2")
)


ggplot(tidy_data, aes(x = body_mass_g,
                     color = species,
                     fill = species)) +
  geom_density(alpha = 0.5) + 
  my_colors


```

## Facets

Faceting is the process of subdividing our data based on one or more categorical variables when plotting. 


```{r}
#| fig-width: 10
#| fig-height: 5
#| warning: false

# an example that involves filtering (via dplyr)
# and facetting to compare conditional distributions

tidy_data |>
  ggplot(aes(x = body_mass_g, fill = species)) +
  geom_histogram() +
  facet_wrap(vars(sex, species), ncol = 3) +
  labs(x = "Body Mass (g)", y = "# Observations",
       title = "Distribution of penguin body mass by sex and species",
       caption = "Data from Palmer Penguins library") +
  my_colors
  
```

## Annotations

### Highlighting via layers

A simple way to highlight data of interest is by layering geoms. 

Here for example we highlight two points in a data set by stacking two `geom_point` layers:


```{r}
#| output: false
# highlight biggest and smallest penguins by body mass
tidy_data |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(
    data = filter(tidy_data, 
                  body_mass_g >= max(body_mass_g) | body_mass_g <= min(body_mass_g)),
    color = "orange",
    size = 3) +  
  geom_point() 
```


### Annotating 


ggplot has an `annotate` function for drawing text, line segments, curves, etc to explicitly call out features in plots.

From the `annotate` help: "This function adds geoms to a plot, but unlike a typical geom function, the properties of the geoms are not mapped from variables of a data frame, but are instead passed in as vectors"

Here's an updated version of the figure above to which we've added text and arrows:


```{r}
tidy_data |>
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(
    data = filter(tidy_data, 
                  body_mass_g >= max(body_mass_g) | body_mass_g <= min(body_mass_g)),
    color = "orange",
    size = 3) +  
  geom_point() +
  annotate(geom = "curve", 
           x = 6000, y = 200, xend = 6300, yend = 220, 
           curvature = .3, arrow = arrow(length = unit(2, "mm"))) +
  annotate(geom = "text", 
           x = 6000, y = 198, 
           label = "Kowalski", hjust = "center") +
  annotate(geom = "curve", 
           x = 2800, y = 210, xend = 2690, yend = 193, 
          curvature = .3, arrow = arrow(length = unit(2, "mm"))) +
  annotate(geom = "text", 
           x = 2800, y = 212, 
           label = "Pingu", hjust = "center")
  
```

## Changing plot limits

If you want to change the limits of a plot, the recommended way to do so is to use the `coord_cartesian` function to specify x and y limits.

```{r}
#| output: false
# zoom in on a particular region of the plot
better_scatter +
  coord_cartesian(xlim = c(4000,5000), ylim=c(180,210))
```

```{r}
#| output: false
# zoom out to include origin
better_scatter +
  coord_cartesian(xlim = c(0,6500), ylim=c(0,240))
```


An alternate functions, `lims` can also be used to specify limits, but `lims`  removes data rather than just changing the limits. This can be important if the plot includes a representation of a statistical function, such a linear model.




## Organizing subplots with patchwork

The [patchwork library](https://patchwork.data-imaginist.com/) is useful for combining multiple plots into a single figure.  

Patchwork uses a simple syntax for specifying the layout of plots, as illustrated below. See the patchwork docs for more details and lots of examples:


```{r}
#| fig-width: 12
#| fig-height: 8
#| warning: false

plot1 <-
  ggplot(tidy_data, 
         aes(x = body_mass_g, 
             fill = species)) + 
  geom_histogram(position = "identity",  # for non-stacked histograms
                 alpha = 0.5) +
  my_colors


plot2 <-
  ggplot(tidy_data, 
         aes(x = flipper_length_mm, 
             fill = species)) + 
  geom_histogram(position = "identity",  
                 alpha = 0.5) +
  my_colors


plot3 <-
  ggplot(tidy_data, 
         aes(x = body_mass_g, 
             y = flipper_length_mm, 
             color = species)) + 
  geom_point() + 
  my_colors

# adding draws plots horizontally
# dividing specifies vertical stacking
fancy_plot <- (plot1 + plot2) / (plot3)

fancy_plot

```


## Exporting a plot

We can export using the `ggsave()` function. By default, `ggsave()` will save the last plot displayed:

```{r}
#| output: false
ggplot(tidy_data, 
       aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point(alpha = 0.5) +
  my_colors

ggsave("bodymass_vs_flipperlength.png", width = 7, height = 5, dpi = 300)
```

If you assign a plot to a variable, you can specify this as the `plot` argument to `ggsave`:


```{r}
#| output: false
ggsave("penguins_multi_plots.pdf", plot = fancy_plot,
       width = 12, height = 8)
```

Note that `ggsave` can infer the appropriate figure format from the filename.